---
title: "RNA Sequencing in Pulmonary Arterial Hypertension"
subtitle: "BMIN503/EPID600 Final Project"
author: "Amber Meservey"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required. I added a new sentence

## Overview {#sec-overview}

Give a brief a description of your project and its goal(s), what data you are using to complete it, and what two faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

Pulmonary arterial hypertension (PAH). is a disease characterized by pathologically increased pressures in the pulmonary vascular bed. Despite recent treatment advances, the disease remains associated with considerable morbidity and mortality. Currently disease diagnosis and prognosis relies simultaneously on invasive hemodynamic measurements and crude clinical associations. Accordingly, there is considerable interest in developing biomarkers for use in diagnosing, distinguishing subtype, treating and following treatment response in (PAH). Our group has previously looked at plasma protein biomarkers in a small clinical trial population of 84 participants with pulmonary arterial hypertension. The proposed project expands on this prior work by examining RNA sequencing in whole blood samples collected from participants at 2 time points during the trial.

The two faculty/staff I've met with regarding this project are Dr. Steven Kawut in Pulmonary Hypertension and Dr. Rui Feng in Biostatistics. Dr. Kawut would be considered my content expert, as he is a master clinician-researcher in PAH, as well as the lead on this clinical trial. He helps to direct appropriate clinical questions being asked of this data. Dr Rui Feng has expertise in RNA sequencing methods as well as statistical methods informing their use. She has been helpful in helping me understand how to format and use appropriate packages for this data.

## Introduction {#sec-introduction}

Describe the problem addressed, its significance, and some background to motivate the problem. This should extend what is in the @sec-overview.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

## Methods {#sec-methods}

Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

```{r}
### Download GEO Query in Bioconductor
if (! requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GEOquery")
```

```{r}
## Load GEO Query
library(GEOquery)
```

```{r}
## Select dataset of interest: PAH dataset, RNA-seq collected from pulmonary smooth artery cells
## ALTERNATIVE: GSE254617
## Download GEO Data: 
data <- getGEO("GSE243193", GSEMatrix = TRUE)
data <- data[[1]]

## View the available phenotypic data:
exprs(data)
sample_metadata <- pData(data)
head(sample_metadata)

## View the expression data as a dataframe
expression_data <- exprs(data)
expression_dataframe <- as.data.frame(expression_data)
```

```{r}

data1 <- getGEO("GSE254617", GSEMatrix = TRUE)
data1 <- data1[[1]]
## View the available phenotypic data:
exprs(data1)
sample_metadata1 <- pData(data1)
head(sample_metadata1)

## View the expression data as a dataframe
expression_data1 <- exprs(data1)
expression_dataframe1 <- as.data.frame(expression_data1)

## Alternative to get raw data: Download data in CSV 
file_path <- "/Users/ambermes/Desktop/GSE254617_txiCounts.csv.gz"
expression_data <- read.csv(gzfile(file_path), header = TRUE, row.names = 1)

file_path_smooth <-"/Users/ambermes/Desktop/GSE243193_PAEC_RNAseq_Counts_GEO2.09.11.23.csv.gz"
expression_data_smooth<- read.csv(gzfile(file_path_smooth), header = TRUE, row.names = 1)

```

\

## Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked here.
